---
title: "Assignment_B1"
author: "IP"
date: "2024-10-22"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(testthat)
library(phyloseq)
library(roxygen2)
```

### Background: abundance data

In my field I work a lot with bacterial abundance data, which is generated as a datatype called a **phyloseq object**. This object is complex but with a highly conserved structure and functionality, and can be converted to a dataframe using the phyloseq::psmelt() function.

The resulting dataframe should always contain columns called "Sample" and "Abundance", alongside columns containing more taxonomic information and metadata (depending on what was originally included by the researcher). Frequently, I want to calculate the *relative* abundance of each bacteria within its sample, which means dividing that bacteria's abundance by the sum of all abundances within that Sample.

Thus I can define a function, calc_relative_abunance(), which takes in a phyloseq object and returns a dataframe containing the psmelt'ed phyloseq object, with relative abundance calculated. As this is a dataframe, this is then very easily used for dplyr manipulation and graphing in ggplot.

The function first checks that my input is a phyloseq object. It then uses psmelt() to convert the phyloseq object into a dataframe, and checks the resulting dataframe contains columns "Sample" and "Abundance" (which the psmelt() function should produce by default).

Once this is confirmed, we filter our data to remove NA Abundance values before grouping by sample and using mutate() to calculate relative abundance.

### Defining the calc_relative_abunance() function

```{r function}
# Relative abundances in phyloseq object
#' @description 
#' Starting from a phyloseq object, this applies the psmelt function to convert data into an R dataframe and subsequently calculate relative abundance of each taxonomic unit within its sample. 
#' 
#' @param ps a whole phyloseq object of taxonomic produced by the phyloseq::phyloseq() function.
#'        ps is named thus as this is common shorthand for a phyloseq object in the field.
#' @return an R dataframe of microbiome composition data containing sample ID and all other 
#' provided metadata, taxonomic information, abundance values, and relative abundance values. 
#' Abundance values of NA are dropped.
#' @examples
#' data(esophagus, package = "phyloseq")
#' data(GlobalPatterns, package = "phyloseq")
#' data(soilrep, package = "phyloseq")
#' calc_relative_abunance(data(esophagus))
#' calc_relative_abunance(data(GlobalPatterns))
#' calc_relative_abunance(data(soilrep))
calc_relative_abunance <- function(ps) {
  
  if(class(ps)!="phyloseq") {
     stop("Input is not a phyloseq object, but rather " , class(ps))}
  
  data <- ps %>%
    phyloseq::psmelt() #convert phyloseq object to an R dataframe
  
  if (!"Sample" %in% colnames(data)) {
    stop("The 'Sample' column is missing from the data.")}
  if (!"Abundance" %in% colnames(data)) {
    stop("The 'Abundance' column is missing from the data.")}
  
   data <- data %>%  
    dplyr::filter(!is.na(Abundance)) %>%
    dplyr::group_by(Sample) %>% 
    dplyr::mutate(Relative_Abundance = Abundance / sum(Abundance)) %>%
    dplyr::ungroup()
  
  return(data)
}
```

### Examples

We are using the soilrep phyloseq object that comes with the phyloseq package. This is passed through our calc_relative_abunance() function, the results of which are piped through head() for easy visualisation as the original dataset is very large.

We can see that a column has been created called "Relative_Abundance".

```{r ex-1}
data(soilrep)
calc_relative_abunance(soilrep) %>% head()
```

In the same way, we can use the GlobalPatterns phyloseq object that also comes with the phyloseq package. This phyloseq object contained much more metadata (eg. sample types, barcodes) and comprehensive taxonomic information (kingdom, phyla, etc.) which is reflected in more columns. However, regardless of the composition of the phyloseq object we still produce the Relative Abundance values!

This demonstrates that our function can take in a phyloseq object of varying complexity and reliably perform the calculation we desire.

```{r ex-2}
data(GlobalPatterns)
calc_relative_abunance(GlobalPatterns) %>% head()
```

In comparison, calling the function with a non-phyloseq input calls our first error message.

```{r error=TRUE}
calc_relative_abunance(1:5) 
calc_relative_abunance(c("This is a character!")) 
```

### Testing the calc_relative_abunance() function

#### Variable specificity

First let's demonstrate that the code calls an error if we don't use a phyloseq object. I have defined a simple dataframe, called df, as part of this test. We expect that any non-phyloseq input returns an error, which is what we see below!

```{r error=TRUE}
df <- data_frame("Num"=1:5,"Let"=c("a","b","c","d","e"))
test_that("Testing variable class specificity", {
  expect_error(calc_relative_abunance(2:10))
  expect_error(calc_relative_abunance("This is a character!"))
  expect_error(calc_relative_abunance(df))
  })
```

#### Correct output

This is a bit harder, as it requires me to put in a phyloseq object for which I already know the desired Relative Abundances. I'll do that in the first chunk, bear with!

```{r make-physeq}
# Create OTU table (matrix of abundance counts)
otu_matrix <- matrix(c(10, 1, 0, 2, 10, 22), nrow=3, byrow=TRUE)
rownames(otu_matrix) <- c("OTU1", "OTU2", "OTU3")
colnames(otu_matrix) <- c("Sample1", "Sample2")
otu_table <- otu_table(otu_matrix, taxa_are_rows=TRUE)

# Create sample data (sample metadata, ie. samples and treatments)
sample_data_frame <- data.frame(SampleID = c("Sample1", "Sample2"),
                                Condition = c("Control", "Treated"))
rownames(sample_data_frame) <- sample_data_frame$SampleID
sample_data <- sample_data(sample_data_frame)

# Create taxonomy table (matrix of taxonomic classification)
taxonomy_matrix <- matrix(c("Bacteria", "Firmicutes", "Clostridia", 
                            "Bacteria", "Proteobacteria", "Gammaproteobacteria",
                            "Bacteria", "Actinobacteria", "Actinobacteria"), 
                          nrow=3, byrow=TRUE)
rownames(taxonomy_matrix) <- c("OTU1", "OTU2", "OTU3")
colnames(taxonomy_matrix) <- c("Domain", "Phylum", "Class")
tax_table <- tax_table(taxonomy_matrix)

# Combine all components into a phyloseq object
sample_physeq <- phyloseq(otu_table, sample_data, tax_table)

# Check structure is corrects
sample_physeq
```

In this sample_physeq we have three OTUs (basically individual taxa) and two Samples. I have set the Abundance values as below, so that we know what the relative abundances should be. I also know psmelt() arranges rows by descending abundance, so this is taken into account when designing our expected test outputs. I have included alongside the expected relative abundance, both the expected abundance and OTUs to make sure that our values are also assigned to the correct OTUs.

| Sample   | OTU  | Abundance | Relative Abundance |
|:---------|:----:|:----------|--------------------|
| Sample 1 | OTU1 | 10        | 10/20 = 0.5        |
| Sample 1 | OTU2 | 0         | 0                  |
| Sample 1 | OTU3 | 10        | 10/20 = 0.5        |
| Sample 2 | OTU1 | 1         | 1/25 = 0.04        |
| Sample 2 | OTU2 | 2         | 2/25 = 0.08        |
| Sample 2 | OTU3 | 22        | 22/25 = 0.88       |

We can now run tests on our function to ensure these values are met! 

```{r output-testing}
expected_rel_abundance <- c(0.88,0.5,0.5,0.08,0.04,0)
expected_abundance <- c(22,10,10,2,1,0)
expected_OTU <- c("OTU3","OTU1","OTU3","OTU2","OTU1","OTU2")
test_that("Testing output accuracy", {
  expect_equal(calc_relative_abunance(sample_physeq)$Relative_Abundance,
               expected_rel_abundance)
  expect_equal(calc_relative_abunance(sample_physeq)$Abundance,
               expected_abundance)
  expect_equal(calc_relative_abunance(sample_physeq)$OTU,
               expected_OTU)
  })
```